Depression and suicidality as costly signals of need

### 2021-04-06 Paper.rmd
Tweaking plots and adding refs to plots

Evolutionary theories of depression tend to either ignore suicidality, or take it as evidence that depression is dysfunctional [for reviews of evolutionary theories of depression, see @].

@@

This study utilized a between-subjects pretest-posttest design to examine how five different emotional signals would influence (1) the degree participants believed a fictional character to be in need (*Belief*) and (2) the likelihood they would provide help (*Action*) in four different vignettes, for a total of 20 conditions.

@@

The sole treatment in this study was the signal participants saw, with these signals including: (1) a verbal request without additional signaling, (2) crying, (3) mild-depression, (4) depression, and (5) a suicide attempt.

@@

Our pre-test and post-test measures, *T1 & T2 Belief* and *T1 & T2 Action*, were all measured on a 0-100 point scale, which we rescaled to 0-1. A substantial number of participants rated their beliefs and actions as exactly 0 or exactly 1 at either T1 or T2. OLS linear regression models are not suitable for a closed and bounded distribution with so many values on the boundary because the residuals would not be normally distributed or have constant variance. Beta regression, a possible alternative, is only defined on the open interval, (0, 1), whereas our data had many values on the boundaries (a closed interval). A variant of beta regression, zero-one inflated beta, is inappropriate because it assumes different processes for the generation of data on the boundaries vs. the interior, and we had no reason to believe that our zeros and ones were generated by a different process than the rest of our observations.

We therefore fit models of our two outcome variables in four ways. First, we fit the preregistered OLS models (reported in the SI). Second, we fit fractional regression models, introduced in a seminal paper by @papke1996 for outcomes on [0, 1], which, in brief, involve fitting a quasi-likelihood version of logistic regression with robust standard errors. This method is implemented in the R `frm` package [@ramalho2019]. Third, we fit GLM logistic regression models, computing 95% CI's via bootstrapping, using the R `glmmboot` package created to model outcomes on [0, 1] [@humphrey2020]. Finally, we fit GLM models with the quasibinomial family and logit link using the `glm` function in R stats library. The estimated coefficients, standard errors, and confidence intervals produced by the latter three methods were *very* similar, and often identical. We therefore proceeded to fit and report models using `glm` with the quasibinomilal family and logit link because model objects produced by this function are compatible with numerous other R packages for estimating and plotting effects.

@@

```{r}
nationT0 <- table(signaling2020$sample)
removed <- sum(!signaling2020$CompleteSurvey)
removed_pct <- 100*signif(removed/nrow(signaling2020), 2)
nationT1 <- table(d$sample)
sexT1 <- table(d$Sex)
```

The study was started by N=`r nrow(signaling2020)` participants who clicked the link to Qualtrics (`r nationT0['American']` US and `r nationT0['Indian']` India). After removing participants who did not finish the survey, failed attention checks, or moved through the study at an unrealistic pace (N=`r removed`, `r removed_pct`%), our final sample was N = `r nrow(d)` (`r nationT1['American']` US and `r nationT1['Indian']` India), with `r sexT1['Male']` males and `r sexT1['Female']` females. For summary statistics, see Table \@ref(tab:sumtab).

@@

Within-subjects, the average effect across all signals on *Belief* at T2 compared to T1 was `r inline_ttest(signal_belief_ttest)`, and on *Action* was `r inline_ttest(signal_action_ttest)` (on the original 0-100 point scale).

@@

The effects of the signals varied substantially by vignette. Signals had the smallest effect in the "Thwarted marriage" vignette, where the effect of the *Suicide attempt* signal on *Belief* and *Action* relative to the *Verbal request* signal was only `r thwart_sv_bdiff` and and `r thwart_sv_adiff` points, respectively (on the original 0-100 point scale). Signals had the largest effect in the "Brother-in-law" vignette, where the effect of the *Suicide attempt* signal on *Belief* and *Action* relative to the *Verbal request* signal was `r brother_sv_bdiff` and and `r brother_sv_adiff` points, respectively. See Figures \@ref(fig:beliefvignette) and \@ref(fig:ecdf).

@@

rom_b_crying <- signif(effect_sizes$est[effect_sizes$id=="Crying,Romantic partner" & effect_sizes$Outcome=="Belief"], 2)
```

Effect sizes (Cohen's d), between- and within-subjects, reveals a similar pattern. The largest between-subjects effect was *Suicide attempt* vs. *Verbal request* on *Belief* in the "Brother-in-law" vignette, d = `r bil_b_suicide`, and the smallest was the effect of *Crying* on *Belief* in the "Romantic partner" vignette, d =`r rom_b_crying`.

@@

We also examined the effect of signaling on how participants perceived the signaler's emotional state. To do this, we created four new variables: LowMood (T1 and T2) and Manipulative (T1 and T2). This was done by adding the dummy variables for whether a participant thought the signaler was depressed, distressed or sad for the LowMood variables and by adding the dummy variables for whether they thought the participant was Devious or Jealous for the Manipulative variables. For each original emotion variable, 0 indicated a participant did not select the emotion, while 1 indicated selection.

We then plotted the mean change in the LowMood and Manipulative variables seen in the data in addition to the change seen in 500 bootstrap resamples (see figure 6).

```{r emotionsplot, fig.width=10, fig.height=8, fig.cap="The effect of the signals on participants' inferred *manipulative* and *low mood* emotional states of the victim, from T1 (arrow base) to T2 (arrow head), in each of the four vignettes. Thick arrows: mean change. Thin arrows: means of 500 bootstrap resamples."}
plot_emotions
```

@@

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. Although possible benefits to depressive states are rarely examined, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can be simultaneously aversive and elicit support [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the responses of others to depressed individuals, especially from those the depressed are in conflict with and those with the capability to provide meaningful benefits.

@@

```{r piloteffects, fig.width=10, fig.cap="Pilot study effect of the signals. A. Effect on Belief at T2, controlling for Belief at T1. B: Proportion of participants who inferred the victim was mentally ill."}
pilot_results$effectplot + (pilot_results$mentalplot + theme(axis.text.y = element_blank())) + plot_annotation(tag_levels = 'A')
```

@@

```{r demoeffectsbeliefs, fig.width=9, fig.height=8, fig.cap="Demographic effects on Belief. **Top**: Association of age with T2 Belief, controlling for T1 Belief, sex, and vignette. **Bottom**: Sex interacting with the effect of the signals on T2 Belief controlling for T1 Belief, age, and vignette. Estimated using a generalized linear regression model with the quasibinomial family."}
effectsPlots$m24plotAge / effectsPlots$m24plotSex
```

```{r demoeffectsaction, fig.width=9, fig.height=8, fig.cap="Demographic effects on Action. **Top**: Association of age with T2 Action, controlling for T1 Action, sex, and vignette. **Bottom**: Association of sex with T2 Action controlling for T1 Action, age, and vignette. Estimated using a generalized linear regression model with the quasibinomial family."}
effectsPlots$m25plotAge / effectsPlots$m25plotSex
```

```{r demovignettebelief, fig.width=8, fig.height=8, fig.cap="Associations of age (top) and sex (bottom) with T2 Belief by vignette, controlling for T1 Belief and signal. Estimated using a generalized linear regression model with the quasibinomial family."}
effectsPlots$m18plot / effectsPlots$m13plot
```

>>>>

Evolutionary theories of depression tend to either ignore suicidality, or take it as evidence that depression is dysfunctional. Wakefield, for instance, who has devoted much of his career to arguing that most cases of depression are simply severe but normal sadness and grief, nevertheless takes suicidality to indicate pathology for some cases of depression [e.g., @wakefield2017]. @nettle2004 argues that the costs of depression, including suicidality, outweigh any benefits, and therefore adaptationist models of depression are implausible. @gilbert2006 sees suicidality as an example of a dysregulated escape defense [for reviews of evolutionary theories of depression, see @].

@@

This study utilized a between-subjects pretest-posttest design to examine how four different emotional signals would influence (1) the degree participants believed a fictional character to be in need (*Belief*) and (2) the likelihood they would provide help (*Action*) relative to a simple verbal request (the control condition), in four different vignettes, for a total of 20 conditions. In this design, the outcomes are measured at pre-treatment (T1), participants are then randomized into either a control group or one or more treatment groups (one of the emotional signals), and the outcomes are measured again (T2). Regression models are used to determine the effect of the treatment conditions on the posttest outcome variables, relative to the control condition, controlling for pretest levels of the outcome variables. We also explored within-subjects effects of the signal on outcomes at T2 compared to T1.

@@

The sole randomized treatment in this study was the signal participants saw, with these signals being one of (1) the control condition: a verbal request without additional signaling, (2) crying, (3) mild-depression, (4) depression, and (5) a suicide attempt.

@@

Our pre-test and post-test measures, *T1 & T2 Belief* and *T1 & T2 Action*, were all measured on a 0-100 point scale. A substantial number of participants rated their beliefs and actions as exactly 0 or exactly 100 at either T1 or T2. OLS linear regression models are not suitable for a closed and bounded distribution with so many values on the boundary because the residuals would not be normally distributed or have constant variance. Beta regression, a possible alternative, is only defined on the open interval, (0, 1), whereas our data had many values on the boundaries (a closed interval). A variant of beta regression, zero-one inflated beta, is inappropriate because it assumes different processes for the generation of data on the boundaries vs. the interior, and we had no reason to believe that our zeros and ones were generated by a different process than the rest of our observations.

We therefore fit models of our two outcome variables (which we rescaled to the continuous interval [0, 1]) in four ways. First, we fit the preregistered OLS models (reported in the SI). Second, we fit fractional regression models, introduced in a seminal paper by @papke1996 for continuous outcomes on the interval [0, 1], which, in brief, involve fitting a quasi-likelihood version of logistic regression with robust standard errors. This method is implemented in the R `frm` package [@ramalho2019]. Third, we fit GLM logistic regression models, computing 95% CI's via bootstrapping, using the R `glmmboot` package created to model continuous outcomes on [0, 1] [@humphrey2020]. Finally, we fit GLM models with the quasibinomial family and logit link using the `glm` function in R stats library, which can accept continous outcomes on [0, 1]. The estimated coefficients, standard errors, and confidence intervals produced by the latter three methods were *very* similar, and often identical. We therefore proceeded to fit and report models using `glm` with the quasibinomilal family and logit link because model objects produced by this function are compatible with numerous other R packages for estimating and plotting effects.

@@

```{r}
nationT0 <- table(signaling2020$sample)
removed <- sum(!signaling2020$CompleteSurvey)
removed_pct <- 100*signif(removed/nrow(signaling2020), 2)
nationT1 <- table(d$sample)
sexT1 <- table(d$Sex)
n_cond <- xtabs(~signal+vignette, d)
```

The study was started by N=`r nrow(signaling2020)` participants who clicked the link to Qualtrics (`r nationT0['American']` US and `r nationT0['Indian']` India). After removing participants who did not finish the survey, failed attention checks, or moved through the study at an unrealistic pace (N=`r removed`, `r removed_pct`%), our final sample was N = `r nrow(d)` (`r nationT1['American']` US and `r nationT1['Indian']` India), with `r sexT1['Male']` males and `r sexT1['Female']` females. The median number of participants per condition was `r median(n_cond)` (min=`r min(n_cond)`, max=`r max(n_cond)`). For the number of participants in each condition, see Table \@ref(tab:conditions). For summary statistics, see Table \@ref(tab:sumtab). For the distributions of participants by age, income, and nationality, see Figure \@ref(fig:socioeconomic).

@@

Within-subjects, the average effect across all signals on *Belief* at T2 compared to T1 was `r inline_ttest(signal_belief_ttest)`, and on *Action* was `r inline_ttest(signal_action_ttest)` (on the original 0-100 point scale). For the change in *Belief* and *Action* for, T1 to T2 for all participants in all conditions, see Figure \@ref(fig:rawbeliefactions2d).

@@

The effects of the signals varied substantially by vignette. Signals had the smallest effect in the "Thwarted marriage" vignette, where the effect of the *Suicide attempt* signal on *Belief* and *Action* relative to the *Verbal request* signal was only `r thwart_sv_bdiff` and `r thwart_sv_adiff` points, respectively (on the original 0-100 point scale). Signals had the largest effect in the "Brother-in-law" vignette, where the effect of the *Suicide attempt* signal on *Belief* and *Action* relative to the *Verbal request* signal was `r brother_sv_bdiff` and and `r brother_sv_adiff` points, respectively. See Figures \@ref(fig:beliefvignette) and \@ref(fig:ecdf).

@@

rom_b_crying <- signif(effect_sizes$est[effect_sizes$id=="Crying,Romantic partner" & effect_sizes$Outcome=="Belief"], 2)

bil_a_suicide <- signif(effect_sizes2$effsize[effect_sizes2$id=="Suicide attempt,Brother-in-law" & effect_sizes2$Outcome=="Action"], 2)

bil_a_verbal <- signif(effect_sizes2$effsize[effect_sizes2$id=="Verbal request,Brother-in-law" & effect_sizes2$Outcome=="Action"], 2)

Effect sizes (Cohen's d), between- and within-subjects, reveals a similar pattern. The largest between-subjects effect was *Suicide attempt* vs. *Verbal request* on *Belief* in the "Brother-in-law" vignette, d = `r bil_b_suicide`, and the smallest was the effect of *Crying* on *Belief* in the "Romantic partner" vignette, d =`r rom_b_crying`. The largest within-subjects effect was *Suicide attempt* vs. *Verbal request* on *Action* in the "Brother-in-law" vignette, d = `r bil_a_suicide`, and the smallest was the *negative* effect of *Verbal request* on *Action* in the "Brother-in-law" vignette, d =`r bil_a_verbal`. See Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2).

@@

The proportion of participants inferring each emotion in each vignette at T1 is depicted in Figure \@ref(fig:t1emotionheatmap), after each signal at T2 in Figure \@ref(fig:t2emotionheatmap), and the change in proportions from T1 to T2 in Figure \@ref(fig:diffemotions). The increase in inferred depression in the *Depression* and *Suicide attempt* conditions, but not the *Verbal request* and *Crying* conditions, helps validate these manipulations, as does the increase in inferred suicidality in the *Suicide attempt* condition.

To further explore the effect of signaling on participants' inferences of the signaler's emotional state, we created four new variables: *Low mood* (T1 and T2) was the sum of the binary variables *Depressed*, *Distressed* and *Sad*; and *Manipulative* (T1 and T2) was the sum of the binary variables *Devious* and *Jealous.* The mean change in the *Low mood* and *Manipulative* variables from T1 to T2 for each signal in each vignette reveals a large decrease in inferred *Manipulation* for costly signals, and a large increase in inferred *Low mood* (with the exception of the Thwarted marriage vignette, in which changes are small). See Figure \@ref(fig:emotionsplot)).

```{r emotionsplot, fig.width=10, fig.height=8, fig.cap="The within-subjects effect of the signals on participants' inferred *Manipulative* and *LowMood* emotional states of the victim, from T1 (arrow base) to T2 (arrow head), for each signal in each of the four vignettes. Thick arrows: mean change. Thin arrows: means of 500 bootstrap resamples."}
plot_emotions
```

@@

Overall, women participants were more likely to believe and help the victim than men, and younger participants were more likely to believe and help the victim than older participants. The caveats are that the sex differences in *Belief* only emerged for the costly signals, the sex differences in *Action* were quite small (but applied across all signals), and the sex and age effects in the Indian sample were non-existent or slightly reversed. We found no significant effects of *Income* on *Belief* or *Action* (results not reported). See Figures \@ref(fig:demographicsignal), \@ref(fig:demovignettebelief), and \@ref(fig:demovignetteaction).

```{r demographicsignal, fig.width=11, fig.height=8, fig.cap="Demographic effects on *Belief* and *Action*. **Top**: Association of age with T2 Belief and Action, controlling for T1 Belief and Action, respectively, and sex and vignette. **Bottom, left**: Sex interacting with the effect of the signals on T2 Belief controlling for T1 Belief, age, and vignette. **Bottom, right**: Main effect of sex on T2 Action, controlling for T1 Action and age. Estimated using generalized linear regression models with the quasibinomial family."}
# effectsPlots$m24plotAge / effectsPlots$m24plotSex

(effectsPlots$m24plotAge + ylim(0, 1) + effectsPlots$m25plotAge + ylim(0, 1))  / (effectsPlots$m24plotSex + effectsPlots$m25plotSex + plot_layout(widths = c(5,1)))

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals helping elicit support when the payoffs to faking need are high [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. Although possible benefits to depressive states are rarely examined, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can be simultaneously aversive and elicit support [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the responses of others to depressed individuals, especially from those the depressed are in conflict with and those with the capability to provide meaningful benefits.

@@

```{r piloteffects, fig.width=10, fig.cap="Pilot study effect of the signals. A. Effect of signal on Belief at T2, controlling for Belief at T1. B: Proportion of participants who inferred the victim was mentally ill. Fit with a generalized linear model with the quasibinomial family."}
pilot_results$effectplot + (pilot_results$mentalplot + theme(axis.text.y = element_blank())) + plot_annotation(tag_levels = 'A')
```

@@

```{r demoeffectsaction, fig.width=9, fig.height=8, fig.cap="Demographic effects on Action. **Top**: Association of age with T2 Action, controlling for T1 Action, sex, and vignette. **Bottom**: Association of sex with T2 Action controlling for T1 Action, age, and vignette. Estimated using a generalized linear regression model with the quasibinomial family."}
# effectsPlots$m25plotAge / effectsPlots$m25plotSex
```

```{r demovignettebelief, fig.width=8, fig.height=8, fig.cap="Associations of age (top) and sex (bottom) with T2 Belief by vignette, controlling for T1 Belief and signal. Estimated using a generalized linear regression model with the quasibinomial family."}
(effectsPlots$m18plot / effectsPlots$m13plot)
```
